A highly-available move operation for replicated trees

### 2019-10-31 paper/move-op.tex
paper: minor changes after a read-through of Martin's editing
Some small language/presentation changes here-and-there.

When the user wants to make a change to the tree, they generate an operation and apply it on their local replica.
Every operation is also asynchronously sent over the network to all other replicas, and applied on every remote replica using the same algorithm as for local operations.
The network may deliver operations in any order; thus, the communication can be performed peer-to-peer, and it does not require any central server or consensus protocol.
We only assume that eventually every operation is applied on every replica, provided that network partitions are of a finite duration.

@@

We can now represent the tree as a set of \isa{(parent, meta, child)} triples, denoted in Isabelle as \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set}.
When we have \isa{(p, m, c) {\isasymin} tree}, that means \isa{c} is a child of \isa{p} in the tree, with associated metadata \isa{m}.
Given a \isa{tree}, we can construct a new \isa{tree'} in which the child \isa{c} is moved to a new parent \isa{p}, with associated metadata \isa{m}, as follows:
\begin{isabelle}
tree' = \{(p', m', c') {\isasymin} tree. c' {\isasymnoteq} c\} {\isasymunion} \{(p, m, c)\}
\end{isabelle}
That is, we remove any existing parent-child relationship for \isa{c} from the set \isa{tree}, and then add \isa{\{(p, m, c)\}} to represent the new parent-child relationship.
This expression appears on lines 31 and 36 of Figure~\ref{fig:code}, as we shall explain shortly.

@@

The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a total order.
This comparison occurs on line 49.

Recall that the replica state includes the operation log (line 14), and we use this log to perform the undo-do-redo cycle.
Lines 45--47 handle the case where the log is empty: in this case, we simply perform the operation using \isa{do\_op}, and return the new tree along with a log containing a single \isa{LogMove} record.
If the log is nonempty (line 48), we take \isa{logop} to be the first element of the log, and \isa{ops} to be the rest.
(The hash character in \isa{logop {\isacharhash} ops} is the \emph{cons} operator that adds one element to the head of a list.)
If the timestamp of \isa{logop} is greater than the timestamp of the new operation (line 49) we first undo \isa{logop} with \isa{undo\_op}, then recursively apply the new operation to the remaining log, and finally reapply \isa{logop} with \isa{redo\_op} (line 50).
Otherwise we perform the operation using \isa{do\_op}, and add the corresponding \isa{LogMove} record as the head of the log (line 51).

@@

This approach to determining ordering also easily supports reordering of child nodes within a parent, e.g.\ to implement the ``bring to front'' and ``send to back'' feature of applications like PowerPoint.
To move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence.
Then we perform a move operation in which the parent node is unchanged, but the metadata is changed to this new ID.

@@

The first invariant we prove is that each tree node has either no parent (if it is the root of a tree) or exactly one parent (if it is a non-root node).
We state this theorem in Isabelle as follows, where \isa{apply\_ops\_unique\_parent} is the name we give to this theorem:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}unique{\isacharunderscore}parent{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree{\isacartoucheclose}
\end{isabelle}
\noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.

@@

The second invariant we require is that the graph must not contain any cycles.
In Isabelle:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}acyclic{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}acyclic\ tree{\isacartoucheclose}
\end{isabelle}
\noindent The \isa{acyclic} predicate is defined on lines 59--60 of Figure~\ref{fig:code}, using the \isa{ancestor} relation (the transitive closure of the graph's edges): a graph contains no cycles if no node is an ancestor of itself.

Proof sketch: Assume \isa{do\_op (op1, tree1) = (op2, tree2)} for some \isa{op1}, \isa{op2}, \isa{tree1} and \isa{tree2}.
Further assume \isa{acyclic tree1}.
We can then prove by contradiction that \isa{tree2} is also acyclic.
Assume to the contrary that \isa{tree2} contains a cycle.
If \isa{do\_op} left the tree unchanged (line 30), we immediately have a contradiction.
Otherwise we can assume \isa{{\isasymnot}(ancestor tree c newp {\isasymor} c = newp)} and we have:
\begin{isabelle}
tree2 = \{(p', m', c') {\isasymin} tree1. c' {\isasymnoteq} c\} {\isasymunion} \{(newp, m, c)\}
\end{isabelle}
That is, \isa{do\_op} adds one edge \isa{\{(newp, m, c)\}} to the graph, and perhaps removes another edge.
Since \isa{tree1} is acyclic, the added edge must lie on the cycle in \isa{tree2}.
Moreover, since only one edge from \isa{newp} to \isa{c} was added, \isa{tree1} must contain a path from \isa{c} to \isa{newp} (which is then closed by the new edge to form a cycle), i.e.\ \isa{c} must be an ancestor of \isa{newp} in \isa{tree1}.
However, in that case, the predicate \isa{ancestor tree1 c newp} on line 30 would have been true, which contradicts the earlier assumption.
Hence, the \isa{do\_op} function preserves the acyclicity invariant of the tree.

@@

As discussed in \S\ref{sec:applying}, we require that when replicas apply the same set of operations, they converge towards the same state, regardless of the order in which the operations are applied.
We formalise this in Isabelle as follows:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}commutes{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}set\ ops{\isadigit{1}}\ {\isacharequal}\ set\ ops{\isadigit{2}}{\isacartoucheclose}\isanewline
\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{1}}{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops{\isadigit{1}}\ {\isacharequal}\ apply{\isacharunderscore}ops\ ops{\isadigit{2}}{\isacartoucheclose}
\end{isabelle}

@@

For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{cons} operator).
We now need to show that
\begin{isabelle}
apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree)) =\\
apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree)).
\end{isabelle}
Let \isa{t3 = log\_time logop} be the timestamp of the operation at the head of the log.
We now consider three cases:
\begin{enumerate}
    \item \isa{t3 < t1 < t2}.
        Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
        Thus, the two operations are applied in the same way as in the base case.
    \item \isa{t1 < t3 < t2}.
        Then \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))} first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
        On the other hand, \isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
        The end result is the same.
    \item \isa{t1 < t2 < t3}.
        Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
        From the inductive hypothesis we have
        \begin{isabelle}
        apply\_op oper2 (apply\_op oper1 (ops, tree')) =\\
        apply\_op oper1 (apply\_op oper2 (ops, tree'))
        \end{isabelle}
        We now apply \isa{oper1} and \isa{oper2} as in the base case, and finally reapply \isa{logop} to both sides using \isa{redo\_op}, resulting in the same state on both sides.
\end{enumerate}

@@

We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
Rather than representing the tree as a set of (parent, metadata, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (metadata, parent) pairs, written in Isabelle as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
The \isa{hashable} type class means that keys must have a hashing function.
In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (\S\ref{sec:unique-parent}).
The hash-map implementation is from the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.

@@

\begin{table}
  \caption{Median response times of a RPC request between replicas in different regions.}
  \label{tab:rpc-times}
  \begin{tabular}{r|rrr}
    \toprule
               & California & Ireland & Singapore \\
    \midrule
    California &            & 395 ms  & 491 ms    \\
    Ireland    & 395 ms     &         & 547 ms    \\
    Singapore  & 491 ms     & 547 ms  &           \\
    \bottomrule
\end{tabular}
\end{table}

@@

Rumor~\cite{Guy:1999gy,RumorManual} is the successor to Ficus.
While previous work uses the kernel filesystem interface, Rumor is a userspace process that is invoked periodically by the user or by a daemon; when run, it compares the state of the replicas.
The original version of Rumor was unable to scale beyond 20 replicas, but an extension called Roam~\cite{Ratner:1999fh} allowed better scaling. 
In an attempt to test Rumor's conflict handling we obtained the source code from archive.org~\cite{RumorSource}; however, we were unable to get it running after modest effort.
%The user manual \cite{RumorManual} makes the claim that "Unix directories are an important example of a kind of file for which Rumor automatically resolves conflicts"

>>>>

When the user wants to make a change to the tree, they generate an operation and apply it to their local replica.
Every operation is also asynchronously sent over the network to all other replicas, and applied to every remote replica using the same algorithm as for local operations.
The network may deliver operations in any order; thus, the communication can be performed peer-to-peer, and it does not require any central server or consensus protocol.
We only assume that eventually every operation is applied on every replica, provided that network partitions are of a finite duration.

@@

We can now represent the tree as a set of \isa{(parent, meta, child)} triples, denoted in Isabelle/HOL as \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set}.
When we have \isa{(p, m, c) {\isasymin} tree}, that means \isa{c} is a child of \isa{p} in the tree, with associated metadata \isa{m}.
Given a \isa{tree}, we can construct a new \isa{tree'} in which the child \isa{c} is moved to a new parent \isa{p}, with associated metadata \isa{m}, as follows:
\begin{isabelle}
tree' = \{(p', m', c') {\isasymin} tree. c' {\isasymnoteq} c\} {\isasymunion} \{(p, m, c)\}
\end{isabelle}
That is, we remove any existing parent-child relationship for \isa{c} from the set \isa{tree}, and then add \isa{\{(p, m, c)\}} to represent the new parent-child relationship.
This expression appears on lines 31 and 36 of Figure~\ref{fig:code}, as we shall explain shortly.

@@

The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
This comparison occurs on line 49.

Recall that the replica state includes the operation log (line 14), and we use this log to perform the undo-do-redo cycle.
Lines 45--47 handle the case where the log is empty: in this case, we simply perform the operation using \isa{do\_op}, and return the new tree along with a log containing a single \isa{LogMove} record.
If the log is nonempty (line 48), we take \isa{logop} to be the first element of the log, and \isa{ops} to be the rest.
(The hash character in \isa{logop {\isacharhash} ops} is the \emph{list cons} operator that adds one element to the head of a list.)
If the timestamp of \isa{logop} is greater than the timestamp of the new operation (line 49) we first undo \isa{logop} with \isa{undo\_op}, then recursively apply the new operation to the remaining log, and finally reapply \isa{logop} with \isa{redo\_op} (line 50).
Otherwise we perform the operation using \isa{do\_op}, and add the corresponding \isa{LogMove} record as the head of the log (line 51).

@@

This approach to determining ordering also easily supports reordering of child nodes within a parent, e.g.\ to implement the ``bring to front'' and ``send to back'' feature of applications like Microsoft PowerPoint.
To move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence.
Then we perform a move operation in which the parent node is unchanged, but the metadata is changed to this new ID.

@@

The first invariant we prove is that each tree node has either no parent (if it is the root of a tree) or exactly one parent (if it is a non-root node).
We state this theorem in Isabelle/HOL as follows, where \isa{apply\_ops\_unique\_parent} is the name we give to this theorem:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}unique{\isacharunderscore}parent{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}unique{\isacharunderscore}parent\ tree{\isacartoucheclose}
\end{isabelle}
\noindent That is, we consider any list of operations \isa{ops} and define \isa{(log, tree)} to be the replica state after \isa{ops} have been applied.
We then prove that \isa{unique\_parent tree} holds, where the \isa{unique\_parent} predicate is defined on lines 56--57 of Figure~\ref{fig:code}: whenever the tree contains a triple whose third element is the child node \isa{c}, then the first and second elements of the triple (the parent node and the metadata) are uniquely defined.

@@

The second invariant we require is that the graph must not contain any cycles.
In Isabelle/HOL:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}acyclic{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops\ {\isacharequal}\ {\isacharparenleft}log{\isacharcomma}\ tree{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}acyclic\ tree{\isacartoucheclose}
\end{isabelle}
\noindent The \isa{acyclic} predicate is defined on lines 59--60 of Figure~\ref{fig:code}, using the \isa{ancestor} relation (the transitive closure of the graph's edges): a graph contains no cycles if no node is an ancestor of itself.

\emph{Proof sketch}: Assume \isa{do\_op (op1, tree1) = (op2, tree2)} for some \isa{op1}, \isa{op2}, \isa{tree1} and \isa{tree2}.
Further assume \isa{acyclic tree1}.
We can then prove by contradiction that \isa{tree2} is also acyclic.
Assume to the contrary that \isa{tree2} contains a cycle.
If \isa{do\_op} left the tree unchanged (line 30), we immediately have a contradiction.
Otherwise we can assume \isa{{\isasymnot}(ancestor tree c newp {\isasymor} c = newp)} and we have:
\begin{isabelle}
tree2 = \{(p', m', c') {\isasymin} tree1. c' {\isasymnoteq} c\} {\isasymunion} \{(newp, m, c)\}
\end{isabelle}
That is, \isa{do\_op} adds one edge \isa{\{(newp, m, c)\}} to the graph, and perhaps removes another edge.
Since \isa{tree1} is acyclic, the added edge must lie on the cycle in \isa{tree2}.
Moreover, since only one edge from \isa{newp} to \isa{c} was added, \isa{tree1} must contain a path from \isa{c} to \isa{newp} (which is then closed by the new edge to form a cycle), i.e.\ \isa{c} must be an ancestor of \isa{newp} in \isa{tree1}.
However, in that case, the predicate \isa{ancestor tree1 c newp} on line 30 would have been true, which contradicts the earlier assumption.
Hence, the \isa{do\_op} function preserves the acyclicity invariant of the tree.

@@

As discussed in \S\ref{sec:applying}, we require that when replicas apply the same set of operations, they converge towards the same state, regardless of the order in which the operations are applied.
We formalise this in Isabelle/HOL as follows:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}commutes{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}set\ ops{\isadigit{1}}\ {\isacharequal}\ set\ ops{\isadigit{2}}{\isacartoucheclose}\isanewline
\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{1}}{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops{\isadigit{1}}\ {\isacharequal}\ apply{\isacharunderscore}ops\ ops{\isadigit{2}}{\isacartoucheclose}
\end{isabelle}

@@

For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{list cons} operator).
We now need to show that:
\begin{isabelle}
apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree)) =\\
apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree)).
\end{isabelle}
Let \isa{t3 = log\_time logop} be the timestamp of the operation at the head of the log.
We now consider three cases:
\begin{enumerate}
    \item \isa{t3 < t1 < t2}.
        Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
        Thus, the two operations are applied in the same way as in the base case.
    \item \isa{t1 < t3 < t2}.
        Then \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))} first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
        On the other hand, \isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
        The end result is the same.
    \item \isa{t1 < t2 < t3}.
        Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
        From the inductive hypothesis we have:
        \begin{isabelle}
        apply\_op oper2 (apply\_op oper1 (ops, tree')) =\\
        apply\_op oper1 (apply\_op oper2 (ops, tree'))
        \end{isabelle}
        We now apply \isa{oper1} and \isa{oper2} as in the base case, and finally reapply \isa{logop} to both sides using \isa{redo\_op}, resulting in the same state on both sides.
\end{enumerate}

@@

We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
Rather than representing the tree as a set of (parent, metadata, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (metadata, parent) pairs, written in Isabelle/HOL as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
The \isa{hashable} type class means that keys must have a hashing function.
In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (\S\ref{sec:unique-parent}).
The hash-map implementation is from the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.

@@

\begin{table}
  \caption{Median response times of a RPC request between replicas in different regions.}
  \label{tab:rpc-times}
  \begin{tabular}{r|rrr}
    \toprule
               & California & Ireland & Singapore \\
    \midrule
    California &      ---      & 395 ms  & 491 ms    \\
    Ireland    & 395 ms     &    ---     & 547 ms    \\
    Singapore  & 491 ms     & 547 ms  &     ---      \\
    \bottomrule
\end{tabular}
\end{table}

@@

Rumor~\cite{Guy:1999gy,RumorManual} is the successor to Ficus.
While previous work uses the kernel filesystem interface, Rumor is a userspace process that is invoked periodically by the user or by a daemon; when run, it compares the state of the replicas.
The original version of Rumor was unable to scale beyond 20 replicas, but an extension called Roam~\cite{Ratner:1999fh} allowed better scaling. 
In an attempt to test Rumor's conflict handling we obtained the source code from \emph{archive.org}~\cite{RumorSource}; however, we were unable to get it running after modest effort.
%The user manual \cite{RumorManual} makes the claim that "Unix directories are an important example of a kind of file for which Rumor automatically resolves conflicts"
